ProbabilityBoundsAnalysis.jl:\\ an arithmetic of probabilities and sets of probabilities

### 2021-09-29 paper/paper.tex
adds sigma convolution

An arithmetic of probability distributions has held a long interest among mathematicians and scientists:
\begin{displayquote}
  A question asked of Kolmogorov, \\
  answered for the sum by Marakov, \\
  partially answered by Sklar and Frank (for which the copula was invented),\\
  made algorithmically available by Williamson, \\
  and generalised by others.
\end{displayquote}

\noindent Indeed it was Kolmogorov who originally asked about the result of a sum of two distributions without knowing their joint distribution. This was answered for the sum by Marakov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, and a now essential object used in probabilistic modelling. In his seminal dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures, and Probability Bounds Analysis (PBA) the name of the method.

@@

\begin{align*}
  u[i] &\leq \overline{F}^{-1}(p) & p_{i} \in [p_{i}, p_{i+1} )\\ 
  d[i] &\geq \underline{F}^{-1}(p) & p_{i} \in [p_{i}, p_{i+1})
\end{align*}

That is, in the region $[p_{i}, p_{i+1})$, $u[i]$ is a lower bound on $\overline{F}^{-1}$ and $d[i]$ is an upper bound on $\underline{F}^{-1}$. Note that the definition of the regions $[p_{i}, p_{i+1})$ or $(p_{i}, p_{i+1}]$ doesn't matter, as long as the choice is consistent. In the case that the p-box is a precise distribution and $\overline{F}^{-1}(p) = \underline{F}^{-1}(p)$ gives:  $u[i+1] = d[i]$. Figure \ref{fig:figure2} shows examples of a distribution and a p-box with various outer representations.

@@

    u1 = quantile.(Normal(m.lo, std.lo), ps_u)
    d1 = quantile.(Normal(m.hi, std.lo), ps_d)

    u2 = quantile.(Normal(m.lo, std.hi), ps_u)
    d2 = quantile.(Normal(m.hi, std.hi), ps_d)

    u = min(u1, u2)
    d = max(d1, d2)

    return pbox(u, d, m, std^2, "normal")
  end

@@

How the moment transformations are performed are not described here, however are available in \textit{MomentArithmetic.jl} \cite{ferson2021distribution}. Figure \ref{fig:figure4} shows an example of an operation between a precise distribution and an interval.

@@

Three important two copulas are:

@@

A minimal structure, cdf, random sample and probability measure of a bivariate p-box is:

@@

@@

\subsection{Uniary operations}

\subsection{Binary operations}
\label{sec:pboxBinary}

\section{An uncertain programming language}
\label{sec:additional_faci}

The long term goal of such a framework is to create a fully uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box or other uncertain quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous and tight fashion. In this section we argue why Julia is an ideal target language for such a framework, and discuss some of the remaining theoretical tasks required to make such a goal a reality.

>>>>

An arithmetic of probability distributions has held a long interest among mathematicians and scientists:
\begin{displayquote}
  A question asked of Kolmogorov, \\
  answered for the sum by Makarov, \\
  partially answered by Sklar and Frank (for which the copula was invented),\\
  made algorithmically available by Williamson, \\
  and generalised by others.
\end{displayquote}

\noindent Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, and a now essential object used in probabilistic modelling. In his seminal dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures, and Probability Bounds Analysis (PBA) the name of the method.

@@

\begin{align*}
  u[i] &\leq \overline{F}^{-1}(p) & p_{i} \in [p_{i}, p_{i+1} )\\ 
  d[i] &\geq \underline{F}^{-1}(p) & p_{i} \in (p_{i}, p_{i+1}]
\end{align*}

That is, in the region $[p_{i}, p_{i+1})$, $u[i]$ is a lower bound on $\overline{F}^{-1}$ and $d[i]$ is an upper bound on $\underline{F}^{-1}$ in $(p_{i}, p_{i+1}]$. In the case that the p-box characterises a precise distribution and $\overline{F}^{-1}(p) = \underline{F}^{-1}(p)$ gives:  $u[i+1] = d[i]$. Figure \ref{fig:figure2} shows examples of a distribution and a p-box with various outer representations.

@@

    u1 = quantile.(Normal(m.lo, std.lo), ps_u)
    u2 = quantile.(Normal(m.lo, std.hi), ps_u)
    u3 = quantile.(Normal(m.lo, std.hi), ps_u)
    u4 = quantile.(Normal(m.hi, std.hi), ps_u)

    d1 = quantile.(Normal(m.lo, std.lo), ps_d)
    d2 = quantile.(Normal(m.lo, std.hi), ps_d)
    d3 = quantile.(Normal(m.hi, std.lo), ps_d)
    d4 = quantile.(Normal(m.hi, std.hi), ps_d)

    u = min.(u1, u2, u3, u4)
    d = max.(d1, d2, d3, d4)

    return pbox(u, d, m, std^2, "normal")
  end
\end{lstlisting}

@@

How the moment transformations are performed are not described here, however are available in \textit{MomentArithmetic.jl}\footnote{https://github.com/AnderGray/MomentArithmetic.jl at time of writing} \cite{ferson2021distribution}. Figure \ref{fig:figure4} shows an example of an operation between a precise distribution and an interval.

@@

Three important 2-copulas are:

@@

A minimal structure, cdf, random sample and probability measure of a bivariate p-box is \footnote{https://github.com/AnderGray/BivariateCopulas.jl can be used for 2-copulas}:

@@

  \centering
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox1.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox2.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox3.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox4.pdf}

  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox5.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox6.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox7.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig8/fig8_pbox8.pdf}

  \caption{Top left shows an initial p-box, and others show results of unary operations on this p-box}
  \label{fig:figure8}

@@

In this section we will discuss how p-boxes can be propagated through unary and binary operations in a rigorous way.

\subsection{Unary operations}
Piggybacking off \textit{IntervalArithmetic.jl}\footnote{https://github.com/JuliaIntervals/IntervalArithmetic.jl at time of writing} \cite{david_p_sanders_2021_5519761}, a p-box can be transformed through any unary operation $f: \mathbb{R} \rightarrow \mathbb{R}$ that has  an interval extension, i.e. accepts intervals. For this, we first transform the p-box into a Dempster-Shafer structure \cite{dempster2008upper,shafer1976mathematical}, propagate the structure through the operation with interval arithmetic, and transform it back to a p-box. A Dempster-Shafer structure (also called a belief function) is a finite collection of intervals, called focal elements $X$, with probabilities attached to each interval, called basic mass assignments $m$, which must sum to $1$. An example of a belief function with three focal elements is: $X = \{[0, 2], [0.5, 1.3], [1, 3]\}$ and $m = \{0.25, 0.5, 0.25\}$. Like a p-box, a Dempster-Shafer also bounds a set of probability distributions, using two set based functions called the \textit{belief} and \textit{plausibility}:

\begin{align*}
  \text{bel}(U) &= \sum_{X \subseteq U} m(X),\\ 
  \text{pl}(U) &= \sum_{X \cap U \neq \emptyset} m(X),
\end{align*}

where the sum is taken over focal elements $X$. For example, the belief and plausibility of $U = [0,2.5]$ are $\text{bel}([0,2.5]) = 0.75$ (\{$[0, 2]$, $[0.5, 1.3]$\} are subsets) and $\text{pl}([0, 2.5]) = 1$ (all focal elements intersect). The belief and plausibility serve as lower and upper bounds on the probability measure:

\begin{equation*}
  \text{bel}(U) \leq \mathbb{P}(U) \leq \text{pl}(U) .
\end{equation*}

A p-box is a special case of a belief function, where the focal elements are ordered by $\leq$. Looking at the leftmost p-box in Figure \ref{fig:figure2} (with $4$ steps) it can be seen that the p-box could be represented with $4$ intervals (i.e. each box in the plot is a focal element). The mass assignment
in this case is the difference in cdf level of each box $p_{i+1} - p_{i}$, which gives an equal mass assignment since we discretised the p-box uniformly. Note that for a general belief function, the focal elements do not have to be ordered, as is in the case of the previous example: $X = \{[0, 2], [0.5, 1.3], [1, 3]\}$. The following function converts a p-box into a belief function:

\begin{lstlisting}[language = Julia]
  function split(F :: pbox)

    n = length(F.u)

    focal_el = interval.(F.u, F.d)
    masses   = ones(n) ./n

    return focal_el, masses
  end
\end{lstlisting}

A belief function can be converted to a p-box by accumulating the belief and plausibility (the lower and upper probability):

\begin{align*}
  \underline{F}(x) &= \text{bel}((-\infty, x]),\\ 
  \overline{F}(x) &= \text{pl}((-\infty, x]),
\end{align*}

which numerically corresponds to sorting the bounds of the focal elements, accumulating the basic mass assignments, and finding the probability levels. Assuming that the basic mass assignments are equal, the following code converts a belief function into a p-box:

\begin{lstlisting}[language = Julia]
  function makepbox(focal_el)

    x_lo = getfield.(focal_el, :lo);
    x_hi = getfield.(focal_el, :hi);

    u = sort(x_lo)
    d = sort(x_hi)

    return pbox(u, d)
  end
\end{lstlisting}

The general function where the focal elements have non-equal mass is simple, but slightly lengthy and will not be described here. A function to perform a unary operation on a p-box is thus:

\begin{lstlisting}[language = Julia]
  function unary(F :: pbox, op)

    focal_el, _ = split(F)
    fe_z = op.(focal_el)

    return makepbox(fe_z)
  end
\end{lstlisting}

Figure \ref{fig:figure8} show examples of unary operations on a p-box.

\subsection{Binary operations}
\label{sec:pboxBinary}
Binary operations on p-boxes are based on \textit{convolutions}, which are operations on two functions which produce a third. A convolution often used in signal processing is:

\begin{equation*}
  f_{Z}(z) = \int^{\infty}_{-\infty}f_{X}(z-x)f_{Y}(x)dx .
\end{equation*}

It is well known that the result of above convolution $f_{Z}$ corresponds to the probability density of the sum of two independent random variables with densities $f_{X}$ and $f_{Y}$. P-box arithmetic is based on generalisations of such convolutions specifically designed for computing binary operations on distribution functions, which work with operations other than sum, and for copulas other than independence. These convolutions, which first appeared in the study of Probabilistic Metric Spaces \cite{schweizer2011probabilistic}, can be be used to compute binary operations between two p-boxes when their copula $C$ is precisely known, only it's lower bound is known $\underline{C}$, or even missing completely. Further, Williamson and Downs \cite{williamson1990probabilistic} describe how robust outer approximations of the convolutions can be found, accounting for the representation error of the two input p-boxes, as well as the error from the integration.

\subsubsection{Operations with known $C$} \hfill \break

A binary operation between two random variables $X$ and $Y$ with cdfs $F_{X}$ and $F_{Y}$ and copula $C$ can be computed with the following Lebesgue-Stieltjes integral:

\begin{equation*}
  \sigma_{C,L}(z) = \int_{L\{z\}}dC(F_{X}(x),F_{Y}(y)) ,
\end{equation*}

where the integration domain is the set $L\{z\} = \{(x,y)| x,y \in \mathbb{R}, L(x,y) < x\}$ (the set of all $x$ and $y$ for which $L(x,y) < z$), and where $L$ is some binary operation (e.g. $L(x,y) = x+y$). The above $\sigma_{C,L}$ convolution is written for precise distributions, however can be extended for p-boxes by performing the convolution with some combination of the p-boxes bounds in the style of interval arithmetic:

\begin{eqnarray*}
  + 
  \begin{cases}
    \underline{F_Z} = \underline{F_X} + \underline{F_Y} \vspace{2mm}\\
    \overline{F_Z} = \overline{F_X} + \overline{F_Y}
  \end{cases}
  \\
  -\begin{cases}
    \underline{F_Z} = \underline{F_X} - \overline{F_Y}\vspace{2mm}\\
    \overline{F_Z} = \overline{F_X} - \underline{F_Y}
  \end{cases}
  \\
  \times
  \begin{cases}
    \underline{F_Z} = \underline{F_X} \times \underline{F_Y}\vspace{2mm}\\
    \overline{F_Z} = \overline{F_X} \times \overline{F_Y}
  \end{cases}
  \\
  \div
  \begin{cases}
    \underline{F_Z} = \underline{F_X} \div \overline{F_Y}\vspace{2mm}\\
    \overline{F_Z} = \overline{F_X} \div \underline{F_Y}
  \end{cases}
\end{eqnarray*}

Williamson and Downs provide an algorithm to compute rigorous bounds on the above convolution, but only under independence (when $C = \Pi$). Around the same time, Yager \cite{yager1986arithmetic} described a similar style convolution on two belief functions under the assumption independence, the algorithm of which is very similar to that of Williamson and Downs. We therefore provide a generalised version of Yager's algorithm where any copula $C$ can be specified. Similar to a unary operation, a binary operation between two p-boxes can be performed by first transforming the bivariate p-box into a two dimensional belief function (which has interval boxes as focal elements), evaluating the operation on the focal elements with interval arithmetic, and transforming the resulting belief function (which will be univariate after the operation) back to a p-box. Converting a bivariate p-box into a 2D belief function is similar to that of the 1D case, however this time the mass assignment $m$ will be determined by the copula $C$. For two univariate focal elements $X$ and $Y$ with probability widths $p_{x} = [p_{x1}, p_{x2}]$ and $p_{y} = [p_{y1}, p_{y2}]$ (i.e. their vertical width in the left Figure \ref{fig:figure2}, with their 1D mass assignment being $p_{x2} - p_{x1}$), the 2D mass assignment on the interval box $X \times Y$ will be:

\begin{equation*}
  C(p_{x2}, p_{y2}) - C(p_{x1}, p_{y2}) - C(p_{x2}, p_{y1}) + C(p_{x1}, p_{y1})
\end{equation*}

The following function converts a bivariate p-box into a 2D belief function:

\begin{lstlisting}[language = Julia]
  function split(F :: BivPbox)

    X = F.X; Y = F.Y; C = F.C

    nx = length(X.u); ny = length(Y.u);

    px = range(0, 1, length = nx + 1);
    py = range(0, 1, length = ny + 1);

    f_el_X = interval.(X.u, X.d);
    f_el_Y = interval.(Y.u, Y.d);

    masses = C(px[2:end], py[2:end]) 
            - C(px[1:end-1], py[2:end]) 
            - C(px[2:end], py[1:end-1]) 
            + C(px[1:end-1], py[1:end-1]);

    f_XY   = [IntervalBox(x,y) for x in f_el_X, 
                              y in f_el_Y];

    return f_XY[:], masses[:]
  end
\end{lstlisting}

A robust $\sigma$ convolution is simply thus:

\begin{lstlisting}[language = Julia]
  function sigma(F :: BivPbox, op)

    nx = length(F.X.u); ny = length(F.Y.d);

    # number of steps of output
    num_steps = max(nx, ny);

    # Convert pbox to a belief function
    fe, masses = split(F)

    # Perform operation with interval arithmetic
    fe_z = [op(f.v[1],f.v[2]) for f in fe]

    # Convert back to p-box
    return makepbox(fe_z, masses, num_steps)
  end
\end{lstlisting}

Again the general function for converting a belief function with non-equal masses is required. The 3rd argument \textit{num\_steps} is the desired discretisation of the output p-box, in a condensation step. The interval operation in the $\sigma$ convolution performs the operation on all combinations of focal elements, so for example if each p-box has 200 steps, the resulting p-box will have $200 \times 200 = 40,000$ steps, which must be reduced in an outer approximation. Again, the general algorithm is simple but lengthy and will be omitted.

\subsubsection{Operations with lower bound $\underline{C}$} \hfill \break

\iffalse
\section{An uncertain programming language}
\label{sec:additional_faci}

The long term goal of such a framework is to create a fully uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box or other uncertain quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous and tight fashion. In this section we argue why Julia is an ideal target language for such a framework, and discuss some of the remaining theoretical tasks required to make such a goal a reality. 
\fi
